---
title: "data.table workshop"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## R basics
Why R? 

Statistics.

I have learnt most of what I know of stats by using R and wading in the documentation of different statistical packages.

See also: open source, documentation, big data, *plotting* (see `ggplot2`).

## What kind of statistics?

- Basic models (see `ezANOVA`)
- Advanced regression (see `lme4`, `nlme`, and `glmm`)
- Machine learning (see `mice`, `caret`, `randomForest`, `e1071`, etc.)
- Bayesian inference (MCMC; see [Stan®](http://mc-stan.org/users/interfaces/rstan))

Plus basically everything else available in other languages.

## Translating from other languages
- Assigment is done with and arrow: `x <- 5` or `5 -> x` (you can assign with `=` but this is usually reserved for arguments)
- (Shortcut on macOS: `Option` + `-`)
- Vectors are created by concatenation: `c(1,2,3,4,5)`
- Multiple data types are stored in lists: `niceList <- list(1, "Strings")`
- These can be named and/or nested: `list("Salami Blessing" = 2, "Miracle Crimes" = c(1,2,3), "nestedList" = niceList)`

----

- Data frames are named tables
- Indexing is done with square brackets: `data[1]`, or by name: `data$SalamiBlessing`
- Mathematical operators are: `+`, `-`, `*`,`/`, `^` or `**`, `x %% y`, `x %/% y`
- Logical operators are: `==`, `<`, `<=`, `!=`, `x & y`, `x | y`

## Basic operations
```{r basicOps}
# Loops
for (i in seq(1,10)){
  print(i)
}

# Function definitions
newFunc <- function(arg1 = 0, arg2 = "type"){
  result <- paste(arg1, arg2)
  return(result)
}

# Linear regression forumlas
model1 <- lm(y ~ x)
summary(model1)
```

## Data wrangling with data.table
`data.table` markets itself as a package for large datasets (e.g. 100GB in RAM). Secondarily, it "Offers a natural and flexible syntax, for faster development."

You can install `data.table` just like every other R package: 
`install.packages("data.table")`
And check the documentation with:
`?data.table` (or if you can't remember `apropos("table")`)

There are some other alternatives that provide similar data manipulation, including `dplyr`.

## From the data.table intro vignette
- *concise* and *consistent* syntax irrespective of the set of operations you would like to perform to achieve your end goal.
- performing analysis *fluidly* without the cognitive burden of having to map each operation to a particular function from a set of functions available before to perform the analysis.
- *automatically* optimising operations internally, and very effectively, by knowing precisely the data required for each operation and therefore very fast and memory efficient.

...if you are interested in reducing *programming* and *compute* time tremendously, then this package is for you. The philosophy that *data.table* adheres to makes this possible.

## Simple use cases

### Reading data
R packages can handle almost all data types (.mat, .pickle, etc.), but who doesn’t like the simplicity of a .csv. But, sometimes, you don’t have control over data formats.

The `fread` function is a "Fast and friendly file finagler", which automatically identifies delimited files and imports them.

----

### Transforming data
Most R packages demand your data in long (rather than wide) format, meaning a row entry for each event, even if this leads to redundant information (e.g. participant IDs).

The `melt` (wide to long) and `dcast` (long to wide) functions for `data.table`s are extensions of the corresponding functions from the `reshape2` package.

```{r melt}
meltedData = melt(data, id.vars = c("id", "condition"),
                measure.vars = c("time1", "time2", "time3"))
```

----

### Summarising data
We often want a quick-and-dirty way of aggregating/partitioning data in meaningful ways. The `data.table` syntax lets you do this on the fly, rather than frantically typing buggy lines in front of your supervisor when you’ve forgotten how many missed trials were in a condition.

```{r summarise, echo = TRUE}
data[is.na(response), .N, by = condition]

data[condition == "forceFed", mean(satiety), by = gender]
```

## More from the data.table vignette
- [Syntax](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html)
- [Reference semantics](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reference-semantics.html)

- [Keys](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-keys-fast-subset.html)
- [Secondary and auto-indexing](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-secondary-indices-and-auto-indexing.html)

### Some more complex examples
```{r complexExamples1, echo = TRUE}
# Remove duplicates
cleanDuplicates <- function(data, cutoff = 0.25) {
  library(data.table)
  data <- as.data.table(data)
  
  # Find CV of duplicates
  cv <- data[ , lapply(.SD, function(x) sd(x, na.rm = T)/mean(x, na.rm = T)), by = "id"]
  
  # Get mean of duplicates
  data <- data[ , lapply(.SD, mean, na.rm = T), by = "id"]
  drop <- cv > cutoff
  data[drop] <- NA # Remove values with high CVs
  return(data)
}
```

----

```{r complexExamples2}
# Create summary variables
data[, ':='(
  firstState = state.name[which.min(question.attempt)],
  lastState = state.name[which.max(question.attempt)],
  maxQuestionAttempt = max(question.attempt),
  modeTrap = modeFunc(state.name),
  screenSecs = max(cumulative.secs.on.screen, na.rm = TRUE),
  lessonTime = max(mins.on.lesson)),
  by = .(userID, tutorialAttempt)]
```

----

```{r complexExamples3}
# Merging data.tables
mergedData <- merge(data1, data2, by = c("id", "session"))
```

## Other resources
- [R for begginers](https://cran.r-project.org/doc/contrib/Paradis-rdebuts_en.pdf)
- The `data.table` [GitHub](https://github.com/Rdatatable/data.table/wiki/Getting-started)

